Interactive Visual Analysis of Image-Centric Cohort Study Data

### 2014-06-24 draft.tex
added Bernhards review

%% indicate IEEE Member or Student Member in form indicated below
\author{P. Klemm$^{1}$, S. Oeltze$^{1}$, K. Lawonn$^{1}$, K. Hegenscheid$^{2}$, H. V{\"o}lzke$^{2}$, B. Preim$^{1}$}
\authorfooter{
%% insert punctuation at end of each item
\item
 $^1$Otto-von-Guericke University Magdeburg, Germany.
\item
 $^2$Ernst-Moritz-Arndt-University Greifswald, Germany.
}

@@

%% Abstract section.
\abstract{%
%(The problem).
%
%The data are analyzed using a priori defined hypotheses to find statistically significant correlations between features (\emph{associations}).
%
%Modern cohort studies incorporate medical image data.
%
%
%The statistically driven epidemiological workflow only allows to determine \emph{associations} between image-derived metrics, such as distances extracted from landmarks of the segmentation model.
%\\\\
% ----------- OLD ABSTRACT ---------
% Epidemiological population studies impose information about a set of subjects (a \emph{cohort}) to characterize disease-specific risk factors.
% %
% Cohort studies comprise of heterogenous variables (\emph{features}) describing the medical condition as well as demographic and lifestyle factors and more recently, medical image data.
% %(New Solution).
% We propose an Interactive Visual Analysis (\emph{IVA}) approach that enables epidemiologists to examine both image-based as well as non-image data, e.g., sociodemographic features and attributes derived from the image data.
% %
% This is achieved by combining brushing and linking enabled coordinated information visualization views and interactive 3D shape renderings with epidemiological data representations such as pivot tables and key figures as association measures.
% %(Validation).
% The presented concepts are applied by experts to gather and evaluate hypotheses about the aging process of the lumbar spine.
% %
% %(Results).
% It allows a more flexible comparison between image and non-image data.
% %
% %(Implications).
% The new framework enables hypotheses validation and generation by incorporating human pattern recognition as well as data mining methods.
% %
% Using all reliable information from the image segmentation linked to non-image features aims to unveil \emph{associations} by applying an iterative analysis approach.
% ----------- /OLD ABSTRACT ---------
\add{
Large-scale longitudinal epidemiological studies investigate groups of people with common characteristics or experiences (a cohort) to characterize health and identify disease-specific risk factors.
%
Cohort study data comprise thousands of individuals and hundreds of heterogenuous variables, describing the medical condition as well as demographic and lifestyle factors, and more recently, medical image data.
%
In the standard epidemiological workflow, a hypothesis is given and validated by analyzing a subset of possibly relevant variables over manually defined subgroups of the cohort, e.g., based on age or body weight.
%
A potential impact of the neglected variables remains unnoticed, data-driven group definition is not supported, and capabilities for integrating the image data into the analysis are missing.\\
%
We propose an Interactive Visual Analysis (IVA) approach that enables epidemiologists to rapidly investigate the entire data pool for both, hypothesis validation and generation.
%
Furthermore, we suggest a strategy for incorporating the image data which involves shape-based object detection and the derivation of attributes describing the object shape.
%
The concurrent investigation of image-based and non-image data is realized in a web-based multiple coordinated view system, comprising standard views from information visualization and epidemiological data representations such as pivot tables.
%
To account for the numerous categorical variables, a mosaic plot is integrated.
%
The views are equipped with brushing facilities and augmented by 3D shape renderings of the segmented objects, e.g., each bar in a histogram is overlayed with a mean shape of the associated subgroup of the cohort.
%
We integrate clustering of variables and object shape for data-driven subgroup definition and statistical key figures for measuring the association between variables.
%
We demonstrate the IVA approach by validating and generating hypotheses related to lower back pain and the aging process of the lumbar spine.}
} % end of abstract

@@

Compiling a list of features for tests of statistical resilience based on experience-driven hypotheses leaves out other features in the data which potentially interact with a disease.
%
This also applies to the chosen landmarks used to quantify medical image data information.
%
The standard workflow lacks methods, which identify correlations that the epidemiologists did not consider.
%
Also, only a small subset of variables can be concurrently analyzed, depending on the statistical measure.
\\\\
We propose an Interactive Visual Analysis (\emph{IVA}) approach \cite{Thomas2005} for the combined analysis of image and non-image data.
%
Visual queries and direct feedback of Visual Analytics systems allow for a fast exploration of the data space incorporating many different variables.
%
Intended as an extension to the well-established epidemiological tools it provides a way to rapidly validate hypotheses and to trigger \emph{hypothesis generation} using data mining methods, such as clustering.
%
\emph{Hypothesis generation} gains importance since the number of epidemiological features increases and the focus shifts towards more complex relations involving more than two features.
\\\\
Our contributions are:
\begin{itemize}
	\item an Interactive Visual Analysis workflow for cohort study data to allow both, hypothesis-driven analysis and hypothesis generation--based on a characterization of the standard epidemiological workflow,
	\item visualization techniques, which incorporate both information visualization and 3D rendering of organ shapes as well as combining them with epidemiological graphics and key figures,
	\item highlighting interesting subject groups and feature associations using shape-based clustering and statistical contingency measures,
%	\item an implementation of the presented methods in a web framework based on \texttt{WebGL}, \texttt{D3.js} and \texttt{NodeJS}.
\end{itemize}
% http://wiki.infowiss.net/Gulf_of_execution ?
We applied our approach to a data set compiled to analyze diseases related to the lumbar spine and aim to determine features, which indicate pathological changes.
%
This data set comprises 127 features and 2 sequences of MRI data from 2,333 patients.
%
We implemented the presented methods using modern web technologies, such as \texttt{WebGL}, \texttt{D3.js} and \texttt{NodeJS}, making them easily accessible for the domain experts.
%
This aims to trigger a fast feedback loop between us and the epidemiologists.

@@

\subsection{Epidemiological Workflow} \label{EpidemiologicalWorkflow}
\begin{figure}[htb]
 \centering
 \includegraphics[width=3.0in]{figures/workflow_comparison}
 \caption{\emph{IVA} tools complement parts of the epidemiological workflow instead of replacing them.
 %
 The appropriate combination of statistical and interactive-driven analysis shows promising potential to unveil the information in the data.
 %
 The standard epidemiology workflow consists of four steps (a).
 %
 IVA complements this workflow and is used to elaborate the feature selection (b).
 %
 The iterative red highlighted part is called the \emph{IVA Loop} and is described in more detail in Figure~\ref{fig:InteractionLoop}.
 %(a) shows the standard epidemiological workflow, (b) the \emph{IVA} supported one. The iterative red highlighted part is called the \emph{IVA Loop} and is described in more detail in Figure~\ref{fig:InteractionLoop}.
 %
 }
  \label{fig:WorkflowComparison}
\end{figure}
The diversity of epidemiology is reflected in the different experts who work at cohort studies, ranging from specialized doctors to medical computer scientists with focus on biometrics, and statisticians.
%
Epidemiologists follow a workflow mainly driven by statistic tools to validate hypotheses about disease-specific risk factors.
%
Following Thew et al \cite{Thew2009}, the workflow can be characterized as follows:
%
\begin{enumerate}
	\item A hypothesis is derived from observations made by clinicians in their daily routine.
%
	\item A set of features depicting conditions affected by the hypothesis is compiled accordingly.
%
	\item Confounding features are identified and taken into account (for example using stratification).
%
	\item Statistical methods, such as regression analysis, assess the association of selected features with the investigated disease.
%
\end{enumerate}
The workflow is shown in Figure~\ref{fig:WorkflowComparison} (a) and serves as orientation for our approach although it does not consider image data.
%
We focus, on the potential of image data and attempt to support \emph{hypothesis generation}.

Reproducibility of results is an epidemiological key requirement.
%
It is difficult to achieve, since many physicians are involved when thousands of test persons are examined and interviewed.
%
Thus, both intra- and inter-observer variability needs to be low for all aspects of a cohort study examination.
%
Longitudinal studies require the acquired features to be comparable for evaluation.
%
%If the data acquisition process changes, an information bias is introduced to the data, hampering inference in and between acquisition cycles.
%

In longitudinal cohort studies, grouping subjects using epidemiological features is essential in order to allow per-group risk determination.
%
Grouping depends on the underlying hypothesis.
%
Age for example is divided into groups (e.g. in 20 year steps) when investigating its influence on a condition.
%
These groups strongly depend on the condition of interest and therefore there is no standard for their categorization.
% --- NEW PARAGRAPH

%
To determine whether a subject is prone to be affected by a certain disease, \emph{relative risks} are determined including confidence intervals that indicate the certainty of that feature being a risk factor.
%To determine whether a subject is prone to be affected by a certain disease, \emph{relative risks} are expressed through the evaluation of p-values, which indicate statistical significance.
%
Statistical correlations are prone to \emph{confounding}, meaning that the association of two features is influenced by a third feature, which needs to be isolated.
%
A famous example for a confounder is the association between shoe size and mortality, where it can be observed that people with larger shoe size have a smaller life expectation than people with small shoe size.
%
The shoe size is actually associated with gender, where women have smaller feet than men but also a longer life expectation.
%
\add{An extreme form of confounding is called \emph{Simpson's Paradox} \cite{SimpsonsParadox}, where the confounder reverses the observed effect between two features.
%
%An example is the higher mortality rate of non-insulin dependent diabetes opposed to insulin dependent diabetes.
A famous example is the Berkeley University gender bias case, where there were significantly less women admitted then man.
%
Analysis later showed that women applied mostly for strongly restricted subjects and when only looking at acceptance divided by departments there was actually a small bias toward women.
%
%For example, insulin dependent and non-insulin dependent diabetes shows 
%In a famous example, researchers concluded that a newer treatment for kidney stones was more effective than traditional surgery, but it was later revealed that the newer treatment was more often being used on small kidney stones.
%For example, comparing mortality rate of non-insulin dependent 
%When for example surgeons are ranked regarding diseased patients, without including the risk of the surgeries, the results may be distorted.
}

@@

\subsection{Epidemiological Data} \label{EpidemiologicalData}
% \com{Primary Reviewer Feedback: Another weakness is on the data analysis; while there is information about the data within the text (including 2.2, 4.1, and 5.2) I would like to see a section that focuses on the data and the challenges thereof.
% This chapter needs to be enhanced! Maybe own section!}
%
% \com{Bernhard: In Section 2 reichen 1-2 allgemeine Sätze, die typische Probleme der Daten beschreiben.}
% \begin{itemize}
% 	\item \com{OpenRefine erwaehnen? Wurde von uns ja nicht mehr verwendet}
% 	\item \com{"I was unsure which type of Image data you had - a better description and the analysis of the data would help here"}
% 	\item \com{Export from SPSS to CSV/JSON - "How do you ensure the semantics of the data is carried over (e.g. for ordinal data)"}
% \end{itemize}
\add{Access to epidemiological data is limited and usually controlled by ethics committees.
%
Large studies, such as the SHIP require written proposals reasoning about each requested features and the expected outcome.
%
Since the number of features and modalities is steadily increasing, the epidemiological community is aware of the lack of specialized analysis methods, making an acceptance more likely.
%In our experience, the applying for a large amount of features required for a \underline{IVA} system requires 
%
Due to the high sensibility of the data, all identity revealing information of a person, such as name or place of residence is anonymized.
}
% --- PARAGRAPH ---

Epidemiological data are highly heterogenous and incomplete.
%
Information about medical history and examinations, genetic conditions, geographical data, questionnaire results and image data yield a complex data space for each subject.
%
For ethical or medical reasons some features cannot be gathered for each subject.
%
The most obvious example are women-specific questions about menstrual status or number of born children.
%
Follow-up examinations or questions about conditions like medications taken after a diagnosed disease also yield features only available for a small amount of subjects.
%

Indicators for medical conditions as well as questions about a subject's lifestyle are often \emph{dichotomous}--they have two manifestations (\emph{Yes} or \emph{No}).
%
Dichotomous data can also be derived by aggregating features to yield only two manifestations (e.g. subjects younger or older than 50 years).
%
Medical examinations mostly comprise categorical (e.g. levels of back pain) and continuous values (e.g. age or body size).
%
\add{Analysis of the data is usually carried out by calculating correlations,
which is challenging due to the data type heterogeneity.
%
Parameter correlation can also be associated with confounding, which can not be automatically predicted but rather has to be judged by a domain expert.
%
Sparse populated features are also hard to assess statistically as to few data samples have a higher chance to distort the real underlying distributions.
%Also, correlation can both be of epidemiological interest but 
}

\subsubsection{Image acquisition} \label{ImageAcquisition} Imaging techniques involving ionizing radiation for the subject are not suitable for ethical reasons.
%
Therefore, MRI is the main method for collecting cohort study imaging data.
%
The image quality is a tradeoff between accuracy and affordability \cite{Preim2014}.
%
This often yields image resolutions inferior to those of clinical day-to-day practice, which makes their analysis more challenging.
%
%The equipment used to gather medical image data is not updated or changed during a longitudinal study to ensure comparability in and between acquisition cycles.
The equipment used to gather medical image data is not updated or changed during a longitudinal study to ensure comparability in and between acquisition cycles.

\subsubsection{Image analysis} \label{ImageAnalysis}
\com{Take up huge amount of space!}
%
Decisions have to be made about \emph{comparison} and \emph{quantification} of image data.
%
Segmentation masks representing the voxels of an anatomical structure would be ideal, since many different key figures, e.g., volume, largest diameter or aspect ratio, can be derived from them.
%
Since reliable and efficient segmentation techniques are not available in general, epidemiologists are forced to measure the data by hand, which is a very tedious work with respect to the number of necessary landmarks and the number of subjects.
%
Information derived by landmarks, such as top and bottom point of a vertebra, are by far not as expressive and versatile as segmentation masks describing its whole shape.
%
They are also prone to a high inter-observer variability and are hard to reproduce.
%
This gains even more importance when analyzing multiple time steps.
%
Morphometric information from landmarks comprises thickness, diameter or length of a structure as well as grey value distribution in an area (used for determining the type of tissue).

@@

\subsection{The Study of Health in Pomerania (SHIP)}
After the pioneering Rotterdam study (started in 1990), several MR imaging study initiatives were initiated.
%
They slightly differ in clinical focus, acquired data and epidemiological research questions.
%
Starting in 1997 with a cohort of 4,308 subjects, the \texttt{SHIP}, located in Northern Germany, aims to characterize health and disease in the widest range possible \cite{Volzke2011}.
%
Data is collected without focus on a group of diseases.
%
This allows the data set to be queried regarding many diseases and conditions.
%
Subjects were examined in a 5-year time span, continuously adding new parameters including MRI scans in the last iteration \cite{Hegenscheid2009}.
%
The MRI protocol features a rich number of sequences.
%
A second cohort \texttt{SHIP-Trend} was established in 2008.
%
The protocols for examining the subjects between \texttt{SHIP} and \texttt{SHIP-Trend} remained the same, making them comparable.
%
The overall examination time for each person attending the study is two days.

@@

\add{%Multiple coordinated views are appropriate for analyzing epidemiological data, because it suits Baldonado et al's \emph{Rule of Diversity} \cite{Baldonado2000}.
%
This section describes prior and related work and covers a broad range of visual analysis methods incorporating both image and non-image data.
%
%We divide these section accordingly and also present lessons learned and from each publication.\\
}
%Designing a visualization, which conveys all data aspects equally, is challenging.
% --- PARAGRAPH ---

@@

\paragraph{Interactive Visual Analysis}
% Interactive Visual Analysis - Steffens Paper
The strength of the \emph{IVA} approach described in the next section is its versatility with respect to the application field \cite{Konyha2009}.
%
Oeltze et al. \cite{Oeltze2007} combined a linked view representation of results from a statistical analysis with feature localizations of the tissue perfusion. % with the goal of its evaluation.
%
This was focused on individual patients and not on cohort study data. 
%

@@

%\section{Interactive Visual Analyis in Cohort Study Data}
\section{Image Centric Cohort Study Data in Interactive Visual Analysis Context} \label{Image Centric Cohort Study Data in Interactive Visual Analysis Context}
We described the epidemiological workflow followed by domain experts where reproducibility and statistical integrity are important aspects (recall Subsection~\ref{EpidemiologicalWorkflow}).
%
Figure~\ref{fig:WorkflowComparison} (a) shows this workflow as a consecutive series of steps.
%
%Introducing the \emph{IVA} principle to the epidemiological domain aims to compensate its weaknesses rather than replacing the existing workflow.
Introducing the \emph{IVA} principle to the epidemiological domain aims to compensate the weaknesses of the existing workflow, rather than replacing it.
%
In the current state, the workflow treats the data like a black box.
%
Statistical tests on features associated to a hypothesis yield a value for deciding whether the data supports the hypothesis or not.
%
Features not included in the analysis may potentially support the chosen hypothesis by discriminating the population in the expected way, but are not highlighted in any way.
%
This becomes even more important when the workflow is adapted to the analysis of the medical image data, where
domain experts annotate landmarks tediously to derive measures, such as diameters.
%
This leaves out the majority of information in the image data by abstracting it to single values.
%
It is easily possible that omitted information would heavily influence the result.
%
Considering more complex parts of the data would make those results more trustworthy and also could identify possible anatomical confounders--an epidemiolgical research result in itself.
\\\\
\emph{IVA} tries to illuminate the black box by making the domain experts part of an iterative feature selection process (see Fig.~\ref{fig:WorkflowComparison} (b)).
%\emph{IVA} tries to illuminate the black box by making the domain experts part of an iterative feature selection process, which is shown in Figure~\ref{fig:WorkflowComparison} (b) as part of the epidemiological workflow.
%
It also aims to project back into the hypothesis formulation step to amplify hypothesis generation.
%
This has to be handled with care, since \emph{overfitting} of expectations to the data is an imminent danger \cite{Turkay2013}.

@@

\paragraph{Local Investigation:} refers to the inspection of dependent variables with respect to certain subsets of the image or shape space.
%
For instance, the epidemiologist selects several lumbar spines with a common characteristic in the image or shape space and wishes to inspect the associated dependent variables in an attribute view \cite{Hermann2014}.
%
The selection step requires dedicated interaction techniques for defining a subset.
%
Alternatively, derived shape-related variables opposed in an attribute view or automatic techniques for shape clustering may be employed \cite{Klemm2013VMV}.
%
Clustering algorithms can be used to investigate associations between shape groups and other non-image variables.
%
Analysis of outliers can indicate segmentation errors or a group of subjects sharing a pathology.

>>>>

%% indicate IEEE Member or Student Member in form indicated below
\author{Paul Klemm$^{1}$, Steffen Oeltze-Jafra$^{1}$, Kai Lawonn$^{1}$, Katrin Hegenscheid$^{2}$, Henry V{\"o}lzke$^{2}$, Bernhard Preim$^{1}$}
\authorfooter{
%% insert punctuation at end of each item
\item
 $^1$Otto-von-Guericke University Magdeburg, Germany.
\item
 $^2$Ernst-Moritz-Arndt-University Greifswald, Germany.
}

@@

%% Abstract section.
\abstract{%
%(The problem).
%The data are analyzed using a priori defined hypotheses to find statistically significant correlations between features (\emph{associations}).
%Modern cohort studies incorporate medical image data.
%The statistically driven epidemiological workflow only allows to determine \emph{associations} between image-derived metrics, such as distances extracted from landmarks of the segmentation model.
%\\\\
% ----------- OLD ABSTRACT ---------
% Epidemiological population studies impose information about a set of subjects (a \emph{cohort}) to characterize disease-specific risk factors.
% %
% Cohort studies comprise of heterogenous variables (\emph{features}) describing the medical condition as well as demographic and lifestyle factors and more recently, medical image data.
% %(New Solution).
% We propose an Interactive Visual Analysis (\emph{IVA}) approach that enables epidemiologists to examine both image-based as well as non-image data, e.g., sociodemographic features and attributes derived from the image data.
% %
% This is achieved by combining brushing and linking enabled coordinated information visualization views and interactive 3D shape renderings with epidemiological data representations such as pivot tables and key figures as association measures.
% %(Validation).
% The presented concepts are applied by experts to gather and evaluate hypotheses about the aging process of the lumbar spine.
% %
% %(Results).
% It allows a more flexible comparison between image and non-image data.
% %
% %(Implications).
% The new framework enables hypotheses validation and generation by incorporating human pattern recognition as well as data mining methods.
% %
% Using all reliable information from the image segmentation linked to non-image features aims to unveil \emph{associations} by applying an iterative analysis approach.
% ----------- /OLD ABSTRACT ---------
\com{Epidemiological Features Begriff einfuehren}
\add{
Large-scale longitudinal epidemiological studies investigate groups of people with common characteristics or experiences (a cohort) to characterize health and identify disease-specific risk factors.
Cohort study data comprise thousands of individuals and hundreds of heterogenuous variables, describing the medical condition as well as demographic and lifestyle factors, and more recently, medical image data.
In the standard epidemiological workflow, a hypothesis is given and validated by analyzing a subset of possibly relevant variables over manually defined subgroups of the cohort, e.g., based on age or body weight.
A potential impact of the neglected variables remains unnoticed, data-driven group definition is not supported, and capabilities for integrating the image data into the analysis are missing.\\
We propose an Interactive Visual Analysis (IVA) approach that enables epidemiologists to rapidly investigate the entire data pool for both, hypothesis validation and generation.
Furthermore, we suggest a strategy for incorporating the image data which involves shape-based object detection and the derivation of attributes describing the object shape.
The concurrent investigation of image-based and non-image data is realized in a web-based multiple coordinated view system, comprising standard views from information visualization and epidemiological data representations such as pivot tables.
To account for the numerous categorical variables, a mosaic plot is integrated.
The views are equipped with brushing facilities and augmented by 3D shape renderings of the segmented objects, e.g., each bar in a histogram is overlayed with a mean shape of the associated subgroup of the cohort.
We integrate clustering of variables and object shape for data-driven subgroup definition and statistical key figures for measuring the association between variables.
We demonstrate the IVA approach by validating and generating hypotheses related to lower back pain and the aging process of the lumbar spine.}
} % end of abstract

@@

Compiling a list of features for tests of statistical resilience based on experience-driven hypotheses leaves out other features in the data which potentially interact with a disease.
%
This also applies to the chosen landmarks used to quantify medical image data information.
%
The standard workflow lacks methods, which identify correlations that the epidemiologists did not consider.
%
Also, only a small subset of variables can be concurrently analyzed. %depending on the statistical measure.
\\\\
We propose an Interactive Visual Analysis (\emph{IVA}) approach \cite{Thomas2005} for the combined analysis of image and non-image data.
%
Visual queries and direct feedback of Visual Analytics systems allow for a fast exploration of the data space incorporating many different variables.
%
Intended as an extension to the well-established epidemiological tools it provides a way to rapidly validate hypotheses and to trigger \emph{hypothesis generation} using data mining methods, such as clustering.
%
\emph{Hypothesis generation} gains importance since the number of epidemiological features increases and the focus shifts towards more complex relations involving more than two features.
\\\\
Our contributions are:
\begin{itemize}
	\item an Interactive Visual Analysis workflow for cohort study data to allow both, hypothesis-driven analysis and hypothesis generation--based on a characterization of the standard epidemiological workflow,
	\item visualization techniques, which incorporate both information visualization and 3D rendering of organ shapes as well as combining them with epidemiological graphics and key figures,
	\item highlighting interesting subject groups and feature associations using shape-based clustering and statistical contingency measures,
%	\item an implementation of the presented methods in a web framework based on \texttt{WebGL}, \texttt{D3.js} and \texttt{NodeJS}.
\end{itemize}
% http://wiki.infowiss.net/Gulf_of_execution ?
We applied our approach to a data set compiled to analyze diseases related to the lumbar spine and aim to determine features, which indicate pathological changes.
%
This data set comprises 127 features and 2 sequences of MRI data from 2,333 patients.
%
We implemented the presented methods using modern web technologies, such as \texttt{WebGL}, \texttt{D3.js} and \texttt{NodeJS}, making them easily accessible for the domain experts to enable a fast feedback loop.
%
This aims to trigger a fast feedback loop between us and the epidemiologists.

@@

\subsection{Epidemiological Workflow} \label{EpidemiologicalWorkflow}
\begin{figure}[htb]
 \centering
 \includegraphics[width=3.0in]{figures/workflow_comparison}
 \caption{
 (a) The standard epidemiology workflow consists of four steps.
 %
 (b) \emph{IVA} tools complement parts of this workflow instead of replacing them.
 %
 The combination of statistical and interactive-driven analysis shows promising potential to unveil information in the data.
 %
 %IVA complements this workflow and is used to elaborate the feature selection.
 %
 We call the iterative red highlighted part \emph{IVA Loop}, described in detail in Figure~\ref{fig:InteractionLoop}.
 %(a) shows the standard epidemiological workflow, (b) the \emph{IVA} supported one. The iterative red highlighted part is called the \emph{IVA Loop} and is described in more detail in Figure~\ref{fig:InteractionLoop}.
 %
 }
  \label{fig:WorkflowComparison}
\end{figure}
The diversity of epidemiology is reflected in the different experts who work at cohort studies, ranging from specialized doctors to medical computer scientists with focus on biometrics, and statisticians.
%
Epidemiologists follow a workflow mainly driven by statistic tools to validate hypotheses about disease-specific risk factors.
%
Following Thew et al \cite{Thew2009}, the workflow can be characterized as follows:
%
\begin{enumerate}
	\item A hypothesis is derived from observations made by clinicians in their daily routine.
%
	\item A set of features depicting conditions affected by the hypothesis is compiled accordingly.
%
	\item Confounding features are identified and taken into account (for example using stratification).
%
	\item Statistical methods, such as regression analysis, assess the association of selected features with the investigated disease.
%
\end{enumerate}
The workflow is shown in Figure~\ref{fig:WorkflowComparison} (a) and serves as orientation for our approach although it does not consider image data.
%
We focus, on the potential of image data and attempt to support \emph{hypothesis generation}.

\emph{Reproducibility} of results is an epidemiological key requirement.
%
It is difficult to achieve, since many physicians are involved when thousands of test persons are examined and interviewed.
%
Thus, both intra- and inter-observer variability needs to be low for all aspects of a cohort study examination.
%
Longitudinal studies require the acquired features to be comparable for evaluation.
%
%If the data acquisition process changes, an information bias is introduced to the data, hampering inference in and between acquisition cycles.
%

Grouping subjects using epidemiological features is essential in cohort studies to allow per-group risk determination.
%
Grouping depends on the underlying hypothesis.
%
Age for example is divided into groups (e.g. in 20 year steps) when investigating its influence.
%
These groups strongly depend on the condition of interest and therefore there is no standard for their categorization.
% --- NEW PARAGRAPH

%
To determine whether a subject is prone to be affected by a certain disease, \emph{relative risks} are determined including confidence intervals that indicate the certainty of that feature being a risk factor.
%To determine whether a subject is prone to be affected by a certain disease, \emph{relative risks} are expressed through the evaluation of p-values, which indicate statistical significance.
%
Statistical correlations are prone to \emph{confounding}, meaning that the association of two features is influenced by a third feature, which needs to be isolated.
%
A famous example for a confounder is the association between shoe size and mortality, where it can be observed that people with larger shoe size have a smaller life expectation. %than people with small shoe size.
%
The shoe size is actually associated with gender, where women have smaller feet and a longer life expectation.
%
\add{An extreme form of confounding is called \emph{Simpson's Paradox} \cite{SimpsonsParadox}, where the confounder reverses the observed effect between two features.
%
%An example is the higher mortality rate of non-insulin dependent diabetes opposed to insulin dependent diabetes.
A famous example is the Berkeley University gender bias case, where significantly less women were admitted.
%
Analysis later showed that women applied mostly for strongly restricted subjects and when only looking at acceptance divided by departments there was actually a small bias toward women.
%
%For example, insulin dependent and non-insulin dependent diabetes shows 
%In a famous example, researchers concluded that a newer treatment for kidney stones was more effective than traditional surgery, but it was later revealed that the newer treatment was more often being used on small kidney stones.
%For example, comparing mortality rate of non-insulin dependent 
%When for example surgeons are ranked regarding diseased patients, without including the risk of the surgeries, the results may be distorted.
}

@@

\subsection{Epidemiological Data} \label{EpidemiologicalData}
% \com{Primary Reviewer Feedback: Another weakness is on the data analysis; while there is information about the data within the text (including 2.2, 4.1, and 5.2) I would like to see a section that focuses on the data and the challenges thereof.
% This chapter needs to be enhanced! Maybe own section!}
%
% \com{Bernhard: In Section 2 reichen 1-2 allgemeine Sätze, die typische Probleme der Daten beschreiben.}
% \begin{itemize}
% 	\item \com{OpenRefine erwaehnen? Wurde von uns ja nicht mehr verwendet}
% 	\item \com{"I was unsure which type of Image data you had - a better description and the analysis of the data would help here"}
% 	\item \com{Export from SPSS to CSV/JSON - "How do you ensure the semantics of the data is carried over (e.g. for ordinal data)"}
% \end{itemize}
\add{Access to epidemiological data is limited and usually controlled by ethics committees.
%
Large studies, such as the SHIP, require written proposals reasoning about each requested features and the expected outcome.
%
Since the number of features and modalities is steadily increasing, the epidemiological community is aware of the lack of specialized analysis methods, making an acceptance more likely.
%In our experience, the applying for a large amount of features required for a \underline{IVA} system requires 
%
Due to the high sensibility of the data, all identity-revealing information of a person, such as name or place of residence is anonymized.
}
% --- PARAGRAPH ---

Epidemiological data are strongly heterogenous and incomplete.
%
Information about medical history and examinations, genetic conditions, geographical data, questionnaire results and image data yield a complex data space for each subject.
%
For ethical or medical reasons some features cannot be gathered for each subject, e.g. women-specific questions about menstrual status or number of born children.
%
Follow-up examinations or questions about conditions such as medications taken after a diagnosed disease also yield features only available for a small amount of subjects.
%

Indicators for medical conditions as well as questions about a subject's lifestyle are often \emph{dichotomous}--they have two manifestations (\emph{Yes} or \emph{No}).
%
Dichotomous data can also be derived by aggregating features to yield only two manifestations (e.g. subjects younger or older than 50 years).
%
Medical examinations comprise categorical (e.g. levels of back pain) and continuous values (e.g. age or body size).
%
\add{Analysis of the data is usually carried out by calculating correlations,
which is challenging due to the data type heterogeneity.
%
Parameter correlation can also be associated with confounding, which can not be automatically predicted but rather has to be judged by a domain expert.
%
Sparse populated features are also hard to assess statistically as too few data samples have a higher chance to distort the real underlying distributions.
%Also, correlation can both be of epidemiological interest but 
}

\paragraph{Image Acquisition.} Imaging techniques involving ionizing radiation for the subject are not suitable for ethical reasons.
%
Therefore, MRI is the main method for collecting cohort study imaging data.
%
The image quality is a tradeoff between accuracy and affordability \cite{Preim2014}.
%
This often yields image resolutions inferior to those of clinical day-to-day practice, which makes their analysis more challenging.
%
%The equipment used to gather medical image data is not updated or changed during a longitudinal study to ensure comparability in and between acquisition cycles.
The equipment used to gather medical image data is not updated or changed during a longitudinal study to ensure comparability in and between acquisition cycles.

\paragraph{Image Analysis.}
\com{Take up huge amount of space!}
%
Decisions have to be made about \emph{comparison} and \emph{quantification} of image data.
%
Segmentation masks representing the voxels of an anatomical structure would be ideal, since many different key figures, e.g., volume, largest diameter or aspect ratio, can be derived from them.
%
Since reliable and efficient segmentation techniques are not available in general, epidemiologists are forced to measure the data by hand. %, which is a very tedious work with respect to the number of necessary landmarks and the number of subjects.
%
Information derived by landmarks, such as top and bottom point of a vertebra, are by far not as expressive and versatile as segmentation masks describing its whole shape.
%
They are also prone to a high inter-observer variability.
%
%This gains even more importance when analyzing multiple time steps.
%
Morphometric information from landmarks comprises thickness, diameter or length of a structure as well as grey value distribution in an area (used for determining the type of tissue).

@@

\subsection{The Study of Health in Pomerania (SHIP)}
After the pioneering Rotterdam study (started in 1990), several MR imaging study initiatives were initiated.
%
They slightly differ in clinical focus, acquired data and epidemiological research questions.
%
Starting in 1997 with a cohort of 4,308 subjects, the \texttt{SHIP}, located in Northern Germany, aims to characterize health and disease in the widest range possible \cite{Volzke2011}.
%
Data is collected without focus on a group of diseases.
%
This allows the data set to be queried regarding many diseases and conditions.
%
Subjects were examined in a 5-year time span, continuously adding new parameters including MRI scans in the last iteration \cite{Hegenscheid2009}.
%
%The MRI protocol features many of sequences.
%
A second cohort \texttt{SHIP-Trend} was established in 2008.
%
The protocols for examining the subjects between \texttt{SHIP} and \texttt{SHIP-Trend} remained the same, making them comparable.
%
The overall examination time for each person attending the study is two days.

@@

\add{%Multiple coordinated views are appropriate for analyzing epidemiological data, because it suits Baldonado et al's \emph{Rule of Diversity} \cite{Baldonado2000}.
%
This section describes prior and related work and covers visual analysis methods incorporating both image and non-image data.
%
%We divide these section accordingly and also present lessons learned and from each publication.\\
}
%Designing a visualization, which conveys all data aspects equally, is challenging.
% --- PARAGRAPH ---

@@

\paragraph{Interactive Visual Analysis.}
% Interactive Visual Analysis - Steffens Paper
The strength of the \emph{IVA} approach described in the next section is its versatility with respect to the application field \cite{Konyha2009}.
%
Oeltze et al. \cite{Oeltze2007} combined a linked view representation of results from a statistical analysis with feature localizations of the tissue perfusion. % with the goal of its evaluation.
%
This was focused on individual patients and not on cohort study data. 
%

@@

%\section{Interactive Visual Analyis in Cohort Study Data}
\section{Image Centric Cohort Study Data in Interactive Visual Analysis Context} \label{Image Centric Cohort Study Data in Interactive Visual Analysis Context}
We described the epidemiological workflow and emphasized the reproducibility and statistical integrity (recall Subsection~\ref{EpidemiologicalWorkflow}).
%
%Figure~\ref{fig:WorkflowComparison} (a) shows this workflow as a consecutive series of steps.
%
%Introducing the \emph{IVA} principle to the epidemiological domain aims to compensate its weaknesses rather than replacing the existing workflow.
Introducing the \emph{IVA} principle to the epidemiological domain aims to compensate the weaknesses of the existing workflow, rather than replacing it (recall Fig.~\ref{fig:WorkflowComparison}).
%
In the current state, the workflow treats the data like a black box.
%
Statistical tests on features associated to a hypothesis yield a value for deciding whether the data supports the hypothesis.
%
Features not included in the analysis may potentially support the chosen hypothesis by discriminating the population in the expected way, but are not highlighted.
%
This becomes even more important when the workflow is adapted to the analysis of the medical image data, where
domain experts annotate landmarks tediously to derive measures, such as diameters.
%
This leaves out the majority of information in the image data by abstracting it to single values.
%
%It is easily possible that omitted information would heavily influence the result.
%
Considering more complex parts of the data would make those results more trustworthy and also could identify possible anatomical confounders--an epidemiolgical research result in itself.
\\\\
\emph{IVA} tries to illuminate the black box by making the domain experts part of an iterative feature selection process (see Fig.~\ref{fig:WorkflowComparison} (b)).
%\emph{IVA} tries to illuminate the black box by making the domain experts part of an iterative feature selection process, which is shown in Figure~\ref{fig:WorkflowComparison} (b) as part of the epidemiological workflow.
%
It also aims to project back into the hypothesis formulation step to amplify hypothesis generation.
%
This has to be handled with care, since \emph{overfitting} of expectations to the data is an imminent danger \cite{Turkay2013}.

@@

\paragraph{Local Investigation.} Refers to the inspection of dependent variables with respect to certain subsets of the image or shape space.
%
For instance, the epidemiologist selects several lumbar spines with a common characteristic in the image or shape space and inspects the associated dependent variables in an attribute view \cite{Hermann2014}.
%
The selection step requires dedicated interaction techniques for defining a subset.
%
Alternatively, derived shape-related variables opposed in an attribute view or automatic techniques for shape clustering may be employed \cite{Klemm2013VMV}.
%
Clustering algorithms can be used to investigate associations between shape groups and other non-image variables.
%
Analysis of outliers can indicate segmentation errors or a group of subjects sharing a pathology.
